

```{r Top Matter}
rm(list = ls()); invisible(gc())

require(rgdal)
require(sp)
require(raster)
require(reshape2)
require(abind)
require(purrr)
require(ggplot2)
require(rgeos)

setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/MainDataFrame/Data")
load("a_mats_UNM_basic.rdata"); rm(m_a_eles, m_a_slop, m_a_aspe)
load("cooneyp_beefy.Rdata")


setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
load(file = "Elevation_Results_Concat.rdata"); rm(m_a_bblats_ele, m_a_bblons_ele, m_aB_ele_plengths)
load(file = "MAT_Results_Concat.rdata"); rm(m_a_bblats_MAT, m_a_bblons_MAT, m_aB_MAT_plengths)
load(file = "VarT_Results_Concat.rdata"); rm(m_a_bblats_VarT, m_a_bblons_VarT, m_aB_VarT_plengths)


wdPAM <- "/Users/boterolab1/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data"
setwd(wdPAM); load("LonLat_BirdPAM_raster.rdata")


'%!in%' <- Negate('%in%')
```

# prep the coooney frame so it is less unwieldy
```{r}
mydata <- cooneyp; rm(cooneyp)
mydata <- mydata[, c("uniquePair", "uniquePairId", "sortcol", "sortorder", "Species.1", "Species.2", "Species.1bl", "Species.2bl",
                     "Pair.age..MY.", "Migration", "Body.mass..log.", "Passerine...non.passerine",
                     "n_pam_cells_sp1", "n_pam_cells_sp2", "Range.overlap",
                     "centroid_lon_sp1", "centroid_lon_sp2", "centroid_lat_sp1", "centroid_lat_sp2", "centroid_distance",
                     "min_lat_sp1", "min_lat_sp2", "max_lat_sp1", "max_lat_sp2", "lat_range_sp1", "lat_range_sp2", "lat_ov_range", "lat_ov_perc_smrnge",
                     "min_ele_sp1", "min_ele_sp2", "max_ele_sp1", "max_ele_sp2", "ele_range_sp1", "ele_range_sp2", "ele_ov_range", "ele_ov_perc_smrnge", "mean_ele_sp1", "mean_ele_sp2",  "med_ele_sp1", "med_ele_sp2", 
                     "min_MAT_sp1", "min_MAT_sp2", "max_MAT_sp1", "max_MAT_sp2", "MAT_range_sp1", "MAT_range_sp2", "MAT_ov_range", "MAT_ov_perc_smrnge", "mean_MAT_sp1", "mean_MAT_sp2",  "med_MAT_sp1", "med_MAT_sp2", 
                     "min_MATr_sp1", "min_MATr_sp2", "max_MATr_sp1", "max_MATr_sp2", "MATr_range_sp1", "MATr_range_sp2", "MATr_ov_range", "MATr_ov_perc_smrnge", "mean_MATr_sp1", "mean_MATr_sp2",  "med_MATr_sp1", "med_MATr_sp2", 
                     "min_VarT_sp1", "min_VarT_sp2", "max_VarT_sp1", "max_VarT_sp2", "VarT_range_sp1", "VarT_range_sp2", "VarT_ov_range", "VarT_ov_perc_smrnge", "mean_VarT_sp1", "mean_VarT_sp2",  "med_VarT_sp1", "med_VarT_sp2",
                     "min_pcp_sp1", "min_pcp_sp2", "max_pcp_sp1", "max_pcp_sp2", "pcp_range_sp1", "pcp_range_sp2", "pcp_ov_range", "pcp_ov_perc_smrnge", "mean_pcp_sp1", "mean_pcp_sp2",  "med_pcp_sp1", "med_pcp_sp2",
                     "min_pcpr_sp1", "min_pcpr_sp2", "max_pcpr_sp1", "max_pcpr_sp2", "pcpr_range_sp1", "pcpr_range_sp2", "pcpr_ov_range", "pcpr_ov_perc_smrnge", "mean_pcpr_sp1", "mean_pcpr_sp2",  "med_pcpr_sp1", "med_pcpr_sp2",
                     "min_VarP_sp1", "min_VarP_sp2", "max_VarP_sp1", "max_VarP_sp2", "VarP_range_sp1", "VarP_range_sp2", "VarP_ov_range", "VarP_ov_perc_smrnge", "mean_VarP_sp1", "mean_VarP_sp2",  "med_VarP_sp1", "med_VarP_sp2"
                     )]
```





```{r Create a SIMPLE 1 row per species dataset}
sum(mydata$Species.1bl != rownames(m_aB_ele_mcosts)) # confirm identical n rows and sort order.
sum(mydata$Species.1bl != rownames(m_aB_MAT_mcosts))
sum(mydata$Species.1bl != rownames(m_aB_VarT_mcosts))

# sympatric cells.
#  - identify paths that should be 0-cost b/c sympatry & replace with appropriate data.
#     - find set that is shared pts in range 1 and range 2
#     - note n sympatric cells in range 1
#     - find cost, len, and bb cells that correspond to the lonlat pts in share set and replace with appropriate data
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data")
load("cbPAM.rdata")
m_a_lonlat <- matrix(paste(m_a_lons,m_a_lats,sep=","),dim(m_a_lons)[1],dim(m_a_lons)[2])
m_a_lonlat[m_a_lonlat == "NA,NA"] <- NA

for (i in 1:nrow(m_a_lonlat)) {
  r12 <- cbPAM[, c("Longitude(x)", "Latitude(y)", mydata$Species.1bl[i], mydata$Species.2bl[i])]
  r12 <- matrix(r12[r12[,3] == 1 & r12[,4] == 1,], ncol = 4)
  r12 <- paste(r12[,1], r12[,2], sep = ",")
  m_a_lonlat[i,][m_a_lonlat[i,] %in% r12] <- "sym"
  m_a_lonlat[i,][m_a_lonlat[i,] %!in% r12 & m_a_lonlat[i,] != "sym" & !is.na(m_a_lonlat[i,])] <- "alo"
  # print(i)
}
m_aB_ele_mcosts[m_a_lonlat == "sym"] <- 0
m_aB_MAT_mcosts[m_a_lonlat == "sym"] <- 0
m_aB_VarT_mcosts[m_a_lonlat == "sym"] <- 0
```

# look at cost distributions (for non-sympatric sp only) to inform summary stats
!!! TAKES A WHILE TO RUN, UNCOMMENT ONLY WHEN NEEDED !!!
```{r}
# x <- m_aB_ele_mcosts
# y <- apply(x, 1, min, na.rm = T)
# x <- x[y > 0 & y < Inf,]
# nrow(x)
# 
# pdf("pair_cost_hists_ELEVATION.pdf", width = 7, height = 7)
# for(i in 1:nrow(x)){
# z <- ggplot()+
#   geom_density(aes(x[i,]))+
#   geom_vline(xintercept = mean(x[i,], na.rm = T), color = "black")+
#   geom_vline(xintercept = quantile(x[i,], 0.50, na.rm=T), color = "red")+
#   geom_vline(xintercept = quantile(x[i,], 0.25, na.rm=T), color = "purple")+
#   geom_vline(xintercept = quantile(x[i,], 0.125, na.rm=T), color = "blue")+
#   geom_vline(xintercept = quantile(x[i,], 0, na.rm=T), color = "gray")
# print(z)
# }
# dev.off()
```


# calculate summary fields of interest
```{r}
mydata$ele_c25 <- apply(m_aB_ele_mcosts, 1, quantile, na.rm = T, probs = 0.25); mydata$ele_c25[is.nan(mydata$ele_c25)] <- NA
# mydata$ele_c12.5 <- apply(m_aB_ele_mcosts, 1, quantile, na.rm = T, probs = 0.125); mydata$ele_c50[is.nan(mydata$ele_c12.5)] <- NA
mydata$ele_c0 <- apply(m_aB_ele_mcosts, 1, quantile, na.rm = T, probs = 0); mydata$ele_c0[is.nan(mydata$ele_c0)] <- NA

mydata$MAT_c25 <- apply(m_aB_MAT_mcosts, 1, quantile, na.rm = T, probs = 0.25); mydata$MAT_c25[is.nan(mydata$MAT_c25)] <- NA
# mydata$MAT_c12.5 <- apply(m_aB_MAT_mcosts, 1, quantile, na.rm = T, probs = 0.125); mydata$MAT_c50[is.nan(mydata$MAT_c12.5)] <- NA
mydata$MAT_c0 <- apply(m_aB_MAT_mcosts, 1, quantile, na.rm = T, probs = 0); mydata$MAT_c0[is.nan(mydata$MAT_c0)] <- NA

mydata$VarT_c25 <- apply(m_aB_VarT_mcosts, 1, quantile, na.rm = T, probs = 0.25); mydata$VarT_c25[is.nan(mydata$VarT_c25)] <- NA
# mydata$VarT_c12.5 <- apply(m_aB_VarT_mcosts, 1, quantile, na.rm = T, probs = 0.125); mydata$VarT_c50[is.nan(mydata$VarT_c12.5)] <- NA
mydata$VarT_c0 <- apply(m_aB_VarT_mcosts, 1, quantile, na.rm = T, probs = 0); mydata$VarT_c0[is.nan(mydata$VarT_c0)] <- NA

setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
save(mydata, file = "Species_Barrier_Data_BONES_beefier.rdata")
```


Prune matrices and calculate summaries.
```{r}
# ele -----------------------------------------------------------------------
ele_c25 <- m_aB_ele_mcosts
# ele_c12.5 <- m_aB_ele_mcosts
ele_c0 <- m_aB_ele_mcosts

ele_lon25 <- m_a_lons
# ele_lon12.5 <- m_a_lons
ele_lon0 <- m_a_lons

ele_lat25 <- m_a_lats
# ele_lat12.5 <- m_a_lats
ele_lat0 <- m_a_lats

for(i in 1:nrow(m_aB_ele_mcosts)){
  ele_c25[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c25[i])] <- NA
  # ele_c12.5[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c12.5[i])] <- NA
  ele_c0[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c0[i])] <- NA
  
  ele_lon25[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c25[i])] <- NA
  # ele_lon12.5[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c12.5[i])] <- NA
  ele_lon0[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c0[i])] <- NA
    
  ele_lat25[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c25[i])] <- NA
  # ele_lat12.5[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c12.5[i])] <- NA
  ele_lat0[i,which(m_aB_ele_mcosts[i,] > mydata$ele_c0[i])] <- NA
}

# sum(!is.na(ele_c25)); sum(!is.na(ele_c12.5)); sum(!is.na(ele_c0)) # working as intended


mydata$centroid_lon_sp1_ele_c25 <- NA
mydata$centroid_lat_sp1_ele_c25 <- NA
mydata$boundary_length_sp1_ele_c25 <- NA
# mydata$centroid_lon_sp1_ele_c12.5 <- NA
# mydata$centroid_lat_sp1_ele_c12.5 <- NA
mydata$centroid_lon_sp1_ele_c0 <- NA
mydata$centroid_lat_sp1_ele_c0 <- NA
for(i in 1:nrow(mydata)){
  
  # c25
  coords1 <- data.frame(cbind(lon=ele_lon25[i,], lat = ele_lat25[i,])); coords1 <- coords1[complete.cases(coords1),]
  coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  centroid1 <- rgeos::gCentroid(coords1)
  mydata$centroid_lon_sp1_ele_c25[i] <- centroid1@coords[1]
  mydata$centroid_lat_sp1_ele_c25[i] <- centroid1@coords[2]  
 
  # new new new ------------------------
  mydata$boundary_length_sp1_ele_c25[i] <- mean(raster::pointDistance(coords1, coords1, lonlat = T, allpairs = T))
  print(i)
  # ------------------------------------
  
  # # c12.5
  # coords1 <- data.frame(cbind(lon=ele_lon12.5[i,], lat = ele_lat12.5[i,])); coords1 <- coords1[complete.cases(coords1),]
  # coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  # centroid1 <- rgeos::gCentroid(coords1)
  # mydata$centroid_lon_sp1_ele_c12.5[i] <- centroid1@coords[1]
  # mydata$centroid_lat_sp1_ele_c12.5[i] <- centroid1@coords[2]
  
  # c0
  coords1 <- data.frame(cbind(lon=ele_lon0[i,], lat = ele_lat0[i,])); coords1 <- coords1[complete.cases(coords1),]
  coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  centroid1 <- rgeos::gCentroid(coords1)
  mydata$centroid_lon_sp1_ele_c0[i] <- centroid1@coords[1]
  mydata$centroid_lat_sp1_ele_c0[i] <- centroid1@coords[2]
}

# MAT -----------------------------------------------------------------------
MAT_c25 <- m_aB_MAT_mcosts
# MAT_c12.5 <- m_aB_MAT_mcosts
MAT_c0 <- m_aB_MAT_mcosts

MAT_lon25 <- m_a_lons
# MAT_lon12.5 <- m_a_lons
MAT_lon0 <- m_a_lons

MAT_lat25 <- m_a_lats
# MAT_lat12.5 <- m_a_lats
MAT_lat0 <- m_a_lats

for(i in 1:nrow(m_aB_MAT_mcosts)){
  MAT_c25[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c25[i])] <- NA
  # MAT_c12.5[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c12.5[i])] <- NA
  MAT_c0[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c0[i])] <- NA
  
  MAT_lon25[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c25[i])] <- NA
  # MAT_lon12.5[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c12.5[i])] <- NA
  MAT_lon0[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c0[i])] <- NA
    
  MAT_lat25[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c25[i])] <- NA
  # MAT_lat12.5[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c12.5[i])] <- NA
  MAT_lat0[i,which(m_aB_MAT_mcosts[i,] > mydata$MAT_c0[i])] <- NA
}


mydata$centroid_lon_sp1_MAT_c25 <- NA
mydata$centroid_lat_sp1_MAT_c25 <- NA
mydata$boundary_length_sp1_MAT_c25 <- NA
# mydata$centroid_lon_sp1_MAT_c12.5 <- NA
# mydata$centroid_lat_sp1_MAT_c12.5 <- NA
mydata$centroid_lon_sp1_MAT_c0 <- NA
mydata$centroid_lat_sp1_MAT_c0 <- NA
for(i in 1:nrow(mydata)){
  
  # c25
  coords1 <- data.frame(cbind(lon=MAT_lon25[i,], lat = MAT_lat25[i,])); coords1 <- coords1[complete.cases(coords1),]
  coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  centroid1 <- rgeos::gCentroid(coords1)
  mydata$centroid_lon_sp1_MAT_c25[i] <- centroid1@coords[1]
  mydata$centroid_lat_sp1_MAT_c25[i] <- centroid1@coords[2]  
   
  # new new new ------------------------
  mydata$boundary_length_sp1_MAT_c25[i] <- mean(raster::pointDistance(coords1, coords1, lonlat = T, allpairs = T))
  print(i)
  # ------------------------------------

  # # c12.5
  # coords1 <- data.frame(cbind(lon=MAT_lon12.5[i,], lat = MAT_lat12.5[i,])); coords1 <- coords1[complete.cases(coords1),]
  # coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  # centroid1 <- rgeos::gCentroid(coords1)
  # mydata$centroid_lon_sp1_MAT_c12.5[i] <- centroid1@coords[1]
  # mydata$centroid_lat_sp1_MAT_c12.5[i] <- centroid1@coords[2]
  
  # c0
  coords1 <- data.frame(cbind(lon=MAT_lon0[i,], lat = MAT_lat0[i,])); coords1 <- coords1[complete.cases(coords1),]
  coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  centroid1 <- rgeos::gCentroid(coords1)
  mydata$centroid_lon_sp1_MAT_c0[i] <- centroid1@coords[1]
  mydata$centroid_lat_sp1_MAT_c0[i] <- centroid1@coords[2]
}


# VarT -----------------------------------------------------------------------
VarT_c25 <- m_aB_VarT_mcosts
# VarT_c12.5 <- m_aB_VarT_mcosts
VarT_c0 <- m_aB_VarT_mcosts

VarT_lon25 <- m_a_lons
# VarT_lon12.5 <- m_a_lons
VarT_lon0 <- m_a_lons

VarT_lat25 <- m_a_lats
# VarT_lat12.5 <- m_a_lats
VarT_lat0 <- m_a_lats

for(i in 1:nrow(m_aB_VarT_mcosts)){
  VarT_c25[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c25[i])] <- NA
  # VarT_c12.5[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c12.5[i])] <- NA
  VarT_c0[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c0[i])] <- NA
  
  VarT_lon25[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c25[i])] <- NA
  # VarT_lon12.5[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c12.5[i])] <- NA
  VarT_lon0[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c0[i])] <- NA
    
  VarT_lat25[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c25[i])] <- NA
  # VarT_lat12.5[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c12.5[i])] <- NA
  VarT_lat0[i,which(m_aB_VarT_mcosts[i,] > mydata$VarT_c0[i])] <- NA
}

# sum(!is.na(VarT_c25)); sum(!is.na(VarT_c12.5)); sum(!is.na(VarT_c0)) # working as intended


mydata$centroid_lon_sp1_VarT_c25 <- NA
mydata$centroid_lat_sp1_VarT_c25 <- NA
# mydata$centroid_lon_sp1_VarT_c12.5 <- NA
# mydata$centroid_lat_sp1_VarT_c12.5 <- NA
mydata$centroid_lon_sp1_VarT_c0 <- NA
mydata$centroid_lat_sp1_VarT_c0 <- NA
for(i in 1:nrow(mydata)){
  
  # c25
  coords1 <- data.frame(cbind(lon=VarT_lon25[i,], lat = VarT_lat25[i,])); coords1 <- coords1[complete.cases(coords1),]
  coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  centroid1 <- rgeos::gCentroid(coords1)
  mydata$centroid_lon_sp1_VarT_c25[i] <- centroid1@coords[1]
  mydata$centroid_lat_sp1_VarT_c25[i] <- centroid1@coords[2]  
 
  # # c12.5
  # coords1 <- data.frame(cbind(lon=VarT_lon12.5[i,], lat = VarT_lat12.5[i,])); coords1 <- coords1[complete.cases(coords1),]
  # coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  # centroid1 <- rgeos::gCentroid(coords1)
  # mydata$centroid_lon_sp1_VarT_c12.5[i] <- centroid1@coords[1]
  # mydata$centroid_lat_sp1_VarT_c12.5[i] <- centroid1@coords[2]
  # 
  # c0
  coords1 <- data.frame(cbind(lon=VarT_lon0[i,], lat = VarT_lat0[i,])); coords1 <- coords1[complete.cases(coords1),]
  coordinates(coords1) <- c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster)
  centroid1 <- rgeos::gCentroid(coords1)
  mydata$centroid_lon_sp1_VarT_c0[i] <- centroid1@coords[1]
  mydata$centroid_lat_sp1_VarT_c0[i] <- centroid1@coords[2]
}
```

Save a species-level dataframe as an intermediate, but this will not be our unit of analysis. 
```{r}
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
save(mydata, file = "Species_Barrier_Data_FEB2021.rdata")
```


Create pair level dataframe
```{r}
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
load(file = "Species_Barrier_Data_FEB2021.rdata")

a <- seq(1:(nrow(mydata)/2)) # fix pair ids so loop below will function correctly
b <- seq(1:(nrow(mydata)/2))
d <- c()
for (i in 1:length(a)) {
  d <- c(d, a[i], b[i])
}
mydata$uniquePairId <- d

mypairs <- unique(mydata$uniquePairId)

i <- 1
pairdata <- mydata[which(mydata$uniquePairId == mypairs[i]), ]
newdata <- pairdata[1,]

# Latitude -----------------------------------------------------
newdata$lat_range_pair_mean <- mean(pairdata$lat_range_sp1, na.rm = T)
newdata$lat_mean_pair <- mean(pairdata$centroid_lat_sp1, na.rm = T)
newdata$lon_mean_pair <- mean(pairdata$centroid_lon_sp1, na.rm = T)

# elevation
newdata$ele_range_pair_mean <- mean(pairdata$ele_range_sp1, na.rm = T)
newdata$lat_mean_pair_ele_c25 <- mean(pairdata$centroid_lat_sp1_ele_c25, na.rm = T)
newdata$lat_mean_pair_ele_c0 <- mean(pairdata$centroid_lat_sp1_ele_c0, na.rm = T)
newdata$lon_mean_pair_ele_c25 <- mean(pairdata$centroid_lon_sp1_ele_c25, na.rm = T)
newdata$lon_mean_pair_ele_c0 <- mean(pairdata$centroid_lon_sp1_ele_c0, na.rm = T)

# MAT
# new new new ---------
newdata$tas_range <- mean(pairdata$mean_MATr_sp1, na.rm = T)
newdata$tas_mean <- mean(pairdata$mean_MAT_sp1, na.rm = T)
# ---------------------
newdata$MAT_range_pair_mean <- mean(pairdata$MAT_range_sp1, na.rm = T)
# ADD LINE HERE FOR TAS MEAN. 
newdata$lat_mean_pair_MAT_c25 <- mean(pairdata$centroid_lat_sp1_MAT_c25, na.rm = T)
newdata$lat_mean_pair_MAT_c0 <- mean(pairdata$centroid_lat_sp1_MAT_c0, na.rm = T)
newdata$lon_mean_pair_MAT_c25 <- mean(pairdata$centroid_lon_sp1_MAT_c25, na.rm = T)
newdata$lon_mean_pair_MAT_c0 <- mean(pairdata$centroid_lon_sp1_MAT_c0, na.rm = T)

# VarT
newdata$VarT_range_pair_mean <- mean(pairdata$VarT_range_sp1, na.rm = T)
newdata$lat_mean_pair_VarT_c25 <- mean(pairdata$centroid_lat_sp1_VarT_c25, na.rm = T)
newdata$lat_mean_pair_VarT_c0 <- mean(pairdata$centroid_lat_sp1_VarT_c0, na.rm = T)
newdata$lon_mean_pair_VarT_c25 <- mean(pairdata$centroid_lon_sp1_VarT_c25, na.rm = T)
newdata$lon_mean_pair_VarT_c0 <- mean(pairdata$centroid_lon_sp1_VarT_c0, na.rm = T)

# pcp
# new new new ---------
newdata$pcp_range <- mean(pairdata$mean_pcpr_sp1, na.rm = T)
newdata$pcp_mean <- mean(pairdata$mean_pcp_sp1, na.rm = T)
# ---------------------

newdata$centroid_distance_ele_c25 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_ele_c25[1],pairdata$centroid_lat_sp1_ele_c25[1]), c(pairdata$centroid_lon_sp1_ele_c25[2],pairdata$centroid_lat_sp1_ele_c25[2]), lonlat = T)
newdata$centroid_distance_ele_c0 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_ele_c0[1],pairdata$centroid_lat_sp1_ele_c0[1]), c(pairdata$centroid_lon_sp1_ele_c0[2],pairdata$centroid_lat_sp1_ele_c0[2]), lonlat = T)
newdata$centroid_distance_MAT_c25 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_MAT_c25[1],pairdata$centroid_lat_sp1_MAT_c25[1]), c(pairdata$centroid_lon_sp1_MAT_c25[2],pairdata$centroid_lat_sp1_MAT_c25[2]), lonlat = T)
newdata$centroid_distance_MAT_c0 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_MAT_c0[1],pairdata$centroid_lat_sp1_MAT_c0[1]), c(pairdata$centroid_lon_sp1_MAT_c0[2],pairdata$centroid_lat_sp1_MAT_c0[2]), lonlat = T)
newdata$centroid_distance_VarT_c25 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_VarT_c25[1],pairdata$centroid_lat_sp1_VarT_c25[1]), c(pairdata$centroid_lon_sp1_VarT_c25[2],pairdata$centroid_lat_sp1_VarT_c25[2]), lonlat = T)
newdata$centroid_distance_VarT_c0 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_VarT_c0[1],pairdata$centroid_lat_sp1_VarT_c0[1]), c(pairdata$centroid_lon_sp1_VarT_c0[2],pairdata$centroid_lat_sp1_VarT_c0[2]), lonlat = T)

# boundary length
newdata$boundary_length_ele_c25 <- mean(pairdata$boundary_length_sp1_ele_c25, na.rm = T)
  
# add lon lat for sp2
newdata$centroid_lon_sp2_ele_c25 <-  newdata$centroid_lon_sp1_ele_c25[2]
newdata$centroid_lat_sp2_ele_c25 <-  newdata$centroid_lat_sp1_ele_c25[2]

newdataseed <- newdata
for (i in 2:max(d)) {
  pairdata <- mydata[which(mydata$uniquePairId == mypairs[i]), ]
  newdata <- pairdata[1,]
  
  # Latitude -----------------------------------------------------
  newdata$lat_range_pair_mean <- mean(pairdata$lat_range_sp1, na.rm = T)
  newdata$lat_mean_pair <- mean(pairdata$centroid_lat_sp1, na.rm = T)
  newdata$lon_mean_pair <- mean(pairdata$centroid_lon_sp1, na.rm = T)
  
  # elevation
  newdata$ele_range_pair_mean <- mean(pairdata$ele_range_sp1, na.rm = T)
  newdata$lat_mean_pair_ele_c25 <- mean(pairdata$centroid_lat_sp1_ele_c25, na.rm = T)
  newdata$lat_mean_pair_ele_c0 <- mean(pairdata$centroid_lat_sp1_ele_c0, na.rm = T)
  newdata$lon_mean_pair_ele_c25 <- mean(pairdata$centroid_lon_sp1_ele_c25, na.rm = T)
  newdata$lon_mean_pair_ele_c0 <- mean(pairdata$centroid_lon_sp1_ele_c0, na.rm = T)
  
  # MAT
  # new new new ---------
  newdata$tas_range <- mean(pairdata$mean_MATr_sp1, na.rm = T)
  newdata$tas_mean <- mean(pairdata$mean_MAT_sp1, na.rm = T)
  # ---------------------
  newdata$MAT_range_pair_mean <- mean(pairdata$MAT_range_sp1, na.rm = T)
  # ADD LINE HERE FOR TAS MEAN. 
  newdata$lat_mean_pair_MAT_c25 <- mean(pairdata$centroid_lat_sp1_MAT_c25, na.rm = T)
  newdata$lat_mean_pair_MAT_c0 <- mean(pairdata$centroid_lat_sp1_MAT_c0, na.rm = T)
  newdata$lon_mean_pair_MAT_c25 <- mean(pairdata$centroid_lon_sp1_MAT_c25, na.rm = T)
  newdata$lon_mean_pair_MAT_c0 <- mean(pairdata$centroid_lon_sp1_MAT_c0, na.rm = T)
  
  # VarT
  newdata$VarT_range_pair_mean <- mean(pairdata$VarT_range_sp1, na.rm = T)
  newdata$lat_mean_pair_VarT_c25 <- mean(pairdata$centroid_lat_sp1_VarT_c25, na.rm = T)
  newdata$lat_mean_pair_VarT_c0 <- mean(pairdata$centroid_lat_sp1_VarT_c0, na.rm = T)
  newdata$lon_mean_pair_VarT_c25 <- mean(pairdata$centroid_lon_sp1_VarT_c25, na.rm = T)
  newdata$lon_mean_pair_VarT_c0 <- mean(pairdata$centroid_lon_sp1_VarT_c0, na.rm = T)
  
  # pcp
  # new new new ---------
  newdata$pcp_range <- mean(pairdata$mean_pcpr_sp1, na.rm = T)
  newdata$pcp_mean <- mean(pairdata$mean_pcp_sp1, na.rm = T)
  # ---------------------

  newdata$centroid_distance_ele_c25 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_ele_c25[1],pairdata$centroid_lat_sp1_ele_c25[1]), c(pairdata$centroid_lon_sp1_ele_c25[2],pairdata$centroid_lat_sp1_ele_c25[2]), lonlat = T)
  newdata$centroid_distance_ele_c0 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_ele_c0[1],pairdata$centroid_lat_sp1_ele_c0[1]), c(pairdata$centroid_lon_sp1_ele_c0[2],pairdata$centroid_lat_sp1_ele_c0[2]), lonlat = T)
  newdata$centroid_distance_MAT_c25 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_MAT_c25[1],pairdata$centroid_lat_sp1_MAT_c25[1]), c(pairdata$centroid_lon_sp1_MAT_c25[2],pairdata$centroid_lat_sp1_MAT_c25[2]), lonlat = T)
  newdata$centroid_distance_MAT_c0 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_MAT_c0[1],pairdata$centroid_lat_sp1_MAT_c0[1]), c(pairdata$centroid_lon_sp1_MAT_c0[2],pairdata$centroid_lat_sp1_MAT_c0[2]), lonlat = T)
  newdata$centroid_distance_VarT_c25 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_VarT_c25[1],pairdata$centroid_lat_sp1_VarT_c25[1]), c(pairdata$centroid_lon_sp1_VarT_c25[2],pairdata$centroid_lat_sp1_VarT_c25[2]), lonlat = T)
  newdata$centroid_distance_VarT_c0 <- raster::pointDistance(c(pairdata$centroid_lon_sp1_VarT_c0[1],pairdata$centroid_lat_sp1_VarT_c0[1]), c(pairdata$centroid_lon_sp1_VarT_c0[2],pairdata$centroid_lat_sp1_VarT_c0[2]), lonlat = T)
  
  # boundary length
  newdata$boundary_length_ele_c25 <- mean(pairdata$boundary_length_sp1_ele_c25, na.rm = T)
  
  # add lon lat for sp2
  newdata$centroid_lon_sp2_ele_c25 <-  newdata$centroid_lon_sp1_ele_c25[2]
  newdata$centroid_lat_sp2_ele_c25 <-  newdata$centroid_lat_sp1_ele_c25[2]

  
  newdataseed <- rbind(newdataseed, newdata)
}
mypairdata <- newdataseed
```

Add realm data -- decided to do this by hand instead.
```{r}
# require(rgdal)
# require(sp)
# require(raster)
# require(reshape2)
# require(abind)
# setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data")
# load(file = "LonLat_BirdPAM_raster.rdata"); rasterpam <- LonLat_BirdPAM_raster; rm(LonLat_BirdPAM_raster)
# 
# setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/WWFRegions/Data")
# my_spdf <- readOGR( dsn= paste0(getwd(),"/wwf_terr_ecos"), layer="wwf_terr_ecos", verbose=FALSE) # load ogr
# my_spdf <- spTransform(my_spdf, crs(rasterpam)); crs(my_spdf) # adjust projection
# spdfdata <- my_spdf@data; str(spdfdata) # save the dataframe
# spdfdata <- spdfdata[,c("OBJECTID", "ECO_NAME", "REALM", "BIOME")] # retain interest columns
# spdfdata$ECO_NAME <- as.character(spdfdata$ECO_NAME) # adjust entry text
# spdfdata$REALM <- as.character(spdfdata$REALM)
# new_spdf <- raster::rasterize(x = my_spdf, y = rasterpam, field = "OBJECTID") # rasterize
# 
# mypairdata$Realm1 <- NA
# mypairdata$Realm2 <- NA
# for (i in 1:length(mypairdata$Realm1)) {
#     # mypairdata$Realm1[i] <- extract(x = new_spdf, y = matrix(data = c(mypairdata$centroid_lon_sp1[i], mypairdata$centroid_lat_sp1[i]), nrow = 1)) # full range.
#     # mypairdata$Realm2[i] <- extract(x = new_spdf, y = matrix(data = c(mypairdata$centroid_lon_sp2[i], mypairdata$centroid_lat_sp2[i]), nrow = 1))
#     mypairdata$Realm1[i] <- extract(x = new_spdf, y = matrix(data = c(mypairdata$centroid_lon_sp1_ele_c25[i], mypairdata$centroid_lat_sp1_ele_c25[i]), nrow = 1)) # path subset
#     mypairdata$Realm2[i] <- extract(x = new_spdf, y = matrix(data = c(mypairdata$centroid_lon_sp2_ele_c25[i], mypairdata$centroid_lat_sp2_ele_c25[i]), nrow = 1))
# } # populated
# 
# for (i in 1:nrow(spdfdata)) {
#   mypairdata$Realm1[mypairdata$Realm1 == spdfdata$OBJECTID[i]] <- spdfdata$REALM[i]
#   mypairdata$Realm2[mypairdata$Realm2 == spdfdata$OBJECTID[i]] <- spdfdata$REALM[i]
#   # print(i)
# } # populated
# 
# sum(is.na(mypairdata$Realm2) & is.na(mypairdata$Realm1)) # few are missing both!
# sum(mypairdata$Realm1 != mypairdata$Realm2, na.rm = T)
# unique(mypairdata$Realm1)
```




Save the pair-level dataframe. 
```{r}
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
rownames(mypairdata) <- mypairdata$Species.1
save(mypairdata, file = "Pair_Barrier_Data_FEB2021.rdata")
```


