---
title: "R Notebook"
output: html_notebook
---


Prepare the Workspace
```{r}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, error = FALSE) # rmd options
rm(list = ls()); invisible(gc()) # cleaning
```
<br><br><br><br><br>

Control Block
```{r} 
# f, f, 0.25 looks nice.
allowmigrants <- F # OPTIONS: T, F 
allowsympatry <- F # OPTIONS: T, F
minoverperc <- 0 # remove pairs that do not have thermal overlap (anagenesis)
costvar <- "ele"
```
<br><br><br><br><br>

Packages & Prefs
```{r}
options(scipen = 999) # turn off scientific notation
'%notin%' <- Negate('%in%')
require(ggplot2) # load packages
require(GGally)
require(viridis)
require(caper)
library(dplyr)
library(stringr)
library(maps)
library(ape)
library(EnvStats)
library(forecast)
require(nlme)
require(geodist)
require(letsR)
require(spdep)
require(spatialreg)
require(rnaturalearth)
require(rnaturalearthdata)
require(rgeos)
require(sf)
require(rgdal)
require(raster)
require(lwgeom)
world <- ne_coastline(scale = "medium", returnclass = "sf")
vlog <- function(x){
   log( x + abs(min( x , na.rm = T)) + 1)
}

setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Analyze_Processed_Cluster_Outputs/Data")
exclusion <- read.csv(file = "exclusion_nonsimpatric_nonmigrant.csv")
exclusion$realm1red[is.na(exclusion$realm1red)] <- "NA" # NA is north america not R's NA value. fix.
exclusion$realm2green[is.na(exclusion$realm2green)] <- "NA"
```
<br><br><br><br><br>

Load Main Data
```{r}
# main dataframe ---------------------------------
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
load(file =  "Pair_Barrier_Data_FEB2021.rdata")
mydata <- mypairdata; rm(mypairdata)
rownames(mydata) <- mydata$Species.1
mydatahold <- mydata
```
<br><br><br><br><br>

initial masks
```{r}
# migration ---
if(allowmigrants == F){
  mydata <- mydata[which(mydata$Migration == 1.0),]
}

# patry
if(allowsympatry == F){
  mydata <- mydata[which(mydata[,paste0(costvar, "_c0")] > 0 ),] # doesnt matter if you use ele, mat, vart paths here, will be same answer
}
```

sort
```{r}
mydata$uniquePairId == exclusion$mydata.uniquePairId
x <- mydata[which(mydata$Species.1 == "Apteryx_owenii"),]
mydata <- mydata[order(match(mydata$uniquePairId,exclusion$mydata.uniquePairId)), ]
y <- mydata[which(mydata$Species.1 == "Apteryx_owenii"),]
# sum(y!=x, na.rm = T) # checks.
# head(mydata)
mydata$uniquePairId == exclusion$mydata.uniquePairId
rm(x,y)
```



```{r}
# # Basic range maps for all pairs (no paths)
# wdPAM <- "/Users/boterolab1/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data"
# setwd(wdPAM); load("cbPAM.rdata")
# setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
# pdf("pairmaps2.pdf", width = 19, height = 9.25)
# for (i in 1:nrow(mydata)){
# x <- cbPAM[,c("Longitude(x)","Latitude(y)",mydata$Species.1bl[i])]
# x <- as.data.frame(x[x[,3] == 1,])
# if(ncol(x) == 1) {
#   x <- t(x)
#   colnames(x) <- c("lon", "lat", "pres")
#   x <- as.data.frame(x)
# } else {
#   colnames(x) <- c("lon", "lat", "pres")
# }
# 
# y <- cbPAM[,c("Longitude(x)","Latitude(y)",mydata$Species.2bl[i])]
# y <- as.data.frame(y[y[,3] == 1,])
# if(ncol(y) == 1) {
#   y <- t(y)
#   colnames(y) <- c("lon", "lat", "pres")
#   y <- as.data.frame(y)
# 
# } else {
#   colnames(y) <- c("lon", "lat", "pres")
# }
# z <- ggplot(world)+
#     geom_sf() +
#     geom_point(data = x, aes(y=lat, x=lon), color = "red") +
#     geom_point(data = y, aes(y=lat, x=lon), color = "green") +
#     theme_bw() +
#     ggtitle(i)
# print(z)
# }
# print(i)
# dev.off()
```

```{r}
mydata_exclusion <- cbind(mydata, exclusion)
save(mydata_exclusion, file = "mydata_exclusion.rdata")
mydata$realm1 <- exclusion$realm1red
mydata$realm2 <- exclusion$realm2green
mydata$realm <- paste0(mydata$realm1, mydata$realm2)
table(mydata$realm)
mydata$realm[mydata$realm == "AAIM"]; mydata$realm[mydata$realm == "IMAA"] <-  "AAIM"
mydata$realm[mydata$realm == "ATIM"]; mydata$realm[mydata$realm == "IMAT"] <-  "ATIM"
mydata$realm[mydata$realm == "ATPA"]; mydata$realm[mydata$realm == "PAAT"] <-  "ATPA"
mydata$realm[mydata$realm == "IMNT"]; mydata$realm[mydata$realm == "NTIM"] <-  "IMNT"
mydata$realm[mydata$realm == "NTAA"]; mydata$realm[mydata$realm == "AANT"] <-  "NTAA"
mydata$realm[mydata$realm == "NTNA"]; mydata$realm[mydata$realm == "NANT"] <-  "NTNA"
mydata$realm[mydata$realm == "PANA"]; mydata$realm[mydata$realm == "NAPA"] <-  "PANA"
mydata$realm <- as.factor(mydata$realm); mydata$realm <- relevel(mydata$realm, "NTNT")

mydata$landgap <- as.logical(exclusion$island)

mydata$cosmopolitan <- as.logical(exclusion$cosmopolitan)

mydata$new.old <- as.logical(exclusion$new.old)

rm(exclusion, mydata_exclusion)
```



Impose masks & do calcuations
```{r}
# filter cosmopolitan and new/old world species (there are relatively few after imposing previous masks.)
mydata <- mydata[which(mydata$cosmopolitan == FALSE & mydata$new.old == FALSE),]

# dependent variable: elevational barrier size ---
mydata$cost <- mydata[, paste0(costvar, "_c25")] 

# data filtering -----------------------
#  thermal overlap ---
mydata$MAT_overlap <- mydata[,paste0("MAT", "_ov_perc_smrnge")]
mydata <- mydata[mydata$MAT_overlap > minoverperc,]

# update sort order --------------------
mydata$sortorder <- seq(1:nrow(mydata))

# longitude ----------------------------
mydata$lon <- mydata[,paste0("lon_mean_pair_", costvar, "_c25")]

# latitude ----------------------------
mydata$lat <- mydata[,paste0("lat_mean_pair_", costvar, "_c25")]

# temperature breadth -----------------
mydata$tas_breadth <- mydata$tas_range # mean(mean(sp1 annual tas range -- one value per cell), mean(sp2 annual tas range -- one value per cell))

# mean annual temperature --------------
mydata$tas_position <- mydata$tas_mean # mean(mean(sp1 annual tas mean  -- one value per cell), mean(sp2 annual tas mean  -- one value per cell))

# precipitation breadth --------------- 
mydata$pcp_breadth <- mydata$pcp_range # mean(mean(sp1 annual pcp range  -- one value per cell), mean(sp2 annual pcp range  -- one value per cell))

# precipitation breadth --------------- 
mydata$pcp_position <- mydata$pcp_mean # mean(mean(sp1 annual pcp mean  -- one value per cell), mean(sp2 annual pcp mean  -- one value per cell))

# distance -----------------------------
mydata$distance <- mydata[,paste0("centroid_distance_",costvar,"_c25")]

# mountain mass ------------------------
mtns <- readOGR(dsn="~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/GMBA", layer="GMBA Mountain Inventory_v1.2-World", verbose = FALSE)
wdPAM <- "/Users/boterolab1/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data"
setwd(wdPAM); load("LonLat_BirdPAM_raster.rdata")
mtns <- rasterize(mtns, LonLat_BirdPAM_raster)
mtns@data@values[!is.na(mtns@data@values)] <- 1 # replace mountain IDs with simple coding. 1 for mountain...
mtns@data@values[is.na(mtns@data@values)] <- 0 # ...0 for no mountain

mydata$mtn_mass <- NA
for (i in 1:nrow(mydata)) {
  coords1<-data.frame(lon=mydata$lon[i], lat=mydata$lat[i]); coordinates(coords1)<-c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster) # get coordinates
  z <- extract(mtns, coords1, buffer = raster::pointDistance(c(0,0), c(0,8), lonlat = T))
  mydata$mtn_mass[i] <- sum(z[[1]]) / length(z[[1]])
}

raster::pointDistance(c(0,0), c(0,8), lonlat = T) # corresponds to a radius of just about 8 degrees.
raster::pointDistance(c(75,75), c(75,(75+8)), lonlat = T) # polar circles are bigger, but not that much bigger. so should be OK.
raster::plot(mtns)
raster::plot(world$geometry, add = T)
plotrix::draw.circle(70, 40, 8, nv = 1000, border = "hotpink", lty = 1, lwd = 1)
plotrix::draw.circle(140, -5, 8, nv = 1000, border = "hotpink", lty = 1, lwd = 1)
plotrix::draw.circle(-80, 10, 8, nv = 1000, border = "hotpink", lty = 1, lwd = 1)
plotrix::draw.circle(-70, -50, 8, nv = 1000, border = "hotpink",lty = 1, lwd = 1)

x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[, "mtn_mass"], decreasing = F),], aes(y=lat, x=lon, color = mtn_mass), alpha = 0.9) +
  ggtitle(i)+scale_color_viridis()
print(x)

# water buffering ----------------------
wtr <- read_sf("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/ne_50m_ocean/ne_50m_ocean.shp")
temp <- mtns
values(temp) <- 100
temp <- mask(temp, wtr)
wtr <- mtns
values(wtr) <- 0
wtr[temp == 100] <- 1; rm(temp)
plot(wtr)

mydata$wtrcells <- NA
mydata$wtrmass <- NA
for (i in 1:nrow(mydata)) {
  coords1<-data.frame(lon=mydata$lon[i], lat=mydata$lat[i]); coordinates(coords1)<-c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster) # get coordinates
  z <- extract(wtr, coords1, buffer = raster::pointDistance(c(0,0), c(0,8), lonlat = T))
  mydata$miniplot[i] <- length(z[[1]])
  mydata$wtrcells[i] <- sum(z[[1]])
  mydata$wtrmass[i] <- sum(z[[1]]) / length(z[[1]])
}
mydata$water_buffering <- mydata$wtrmass

x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[, "water_buffering"], decreasing = F),], aes(y=lat, x=lon, color = water_buffering), alpha = 0.9) +
  ggtitle(i)+scale_color_viridis()
print(x)


# dispersal ability --------------------
dispab <- read.csv("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/Bird Hand-Wing Index/Dataset HWI 2020-04-10.csv")
mydata$dispersal_ability <- NA
for (i in 1:nrow(mydata)){
  dispab_sp1 <- dispab$HWI[dispab$IUCN.name == mydata$Species.1bl[i]]
  dispab_sp2 <- dispab$HWI[dispab$IUCN.name == mydata$Species.2bl[i]]
  dispab_pair <- mean(c(dispab_sp1, dispab_sp2), na.rm = T)
  mydata$dispersal_ability[i] <- dispab_pair
  rm(dispab_sp1, dispab_sp2, dispab_pair)
}
mydata <- mydata[!is.nan(mydata$dispersal_ability),]

# pair age -----------------------------
mydata$pair_age <- mydata$Pair.age..MY.

# length of boundary -------------------
mydata$boundary_length <- mydata$boundary_length_ele_c25

# retain cols of interest only.
mydata <- mydata[,c("uniquePairId", "Species.1", "Species.2", "Species.1bl", "Species.2bl", "cost", "lat", "lon",
                    "tas_breadth","tas_position", "pcp_breadth","pcp_position", "mtn_mass", "water_buffering", 
                    "dispersal_ability", "pair_age", "distance", "boundary_length", "MAT_overlap", "realm",
                    "landgap")]
```

load phylo and prune
```{r}
# phylo ------------------------------------------
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/BirdTrees")
load(file = "BirdTrees.Rdata")
tree <- trees[[1]]; rm(trees) # pick tree (VF GET TREES FROM COONEY!!! and use MCC tree.) -- currently just using tree 1 here. 
tree <- drop.tip(tree, tree$tip.label[which(tree$tip.label %notin% mydata$Species.1)]) # initial name matching.
mydata <- mydata[which(mydata$Species.1 %in% tree$tip.label),]
mydata <- mydata[match(tree$tip.label, mydata$Species.1),]
sum(mydata$Species.1 != tree$tip.label) # sorted (but still specify form below for safety)
```

Save freame for CB
```{r}
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Analyze_Processed_Cluster_Outputs/Data")
save(mydata, file = "ele_data_for_CB.rdata")
write.csv(mydata, file = "ele_data_for_CB.csv")
```




UPDATED CALCS FOR MTN MASS 
```{r}
lonlat2UTM = function(lonlat) {
  utm = (floor((lonlat[1] + 180) / 6) %% 60) + 1
  if(lonlat[2] > 0) {
    utm + 32600
  } else{
    utm + 32700
  }
}

mtns <- readOGR(dsn="~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/GMBA", layer="GMBA Mountain Inventory_v1.2-World", verbose = FALSE)
mtns <- st_as_sf(mtns)
buffer_ranges<-c(seq(from = 0, to = 200, by  = 50)[-1], seq(from=0, to = 3000, by = 250)[-1]) * 1000
bufferdata <- matrix(nrow = nrow(mydata), ncol = length(buffer_ranges))
for(i in 1:nrow(mydata)){
  my_centroid <- st_geometry(st_sfc(st_point(c(mydata$lon[i],mydata$lat[i]))))
  st_crs(my_centroid) <- crs(LonLat_BirdPAM_raster)
  EPSG_2_UTM <- as.numeric(lonlat2UTM(my_centroid[[1]]))
  # To see the UTM #st_crs(EPSG_2_UTM)$proj4string
  my_centroid_proj = st_transform(st_as_sf(my_centroid), EPSG_2_UTM)
  for(j in 1:length(buffer_ranges)){
    my_centroid_with_b_km_buffer <- st_buffer(my_centroid_proj,dist=buffer_ranges[j])
    mtns_in_buffer_b_km <- st_intersection(st_make_valid(sf::st_transform(my_centroid_with_b_km_buffer, crs=st_crs(mtns))), st_make_valid(mtns))
    area_in_buffer_b_km <- sum(as.numeric(st_area(mtns_in_buffer_b_km)))
    bufferdata[i, j] <- area_in_buffer_b_km
  }
  print(i)
}
bufferdatahold <- bufferdata
for(i in 1:nrow(bufferdata)) {
  bufferdata[i,] <- bufferdata[i,] / buffer_ranges
}
mydata$mtn_mass_new <- rowMeans(bufferdata)
plot(scale(mydata$mtn_mass_new)~scale(mydata$mtn_mass))
abline(0,1)

x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[, "mtn_mass_new"], decreasing = F),], aes(y=lat, x=lon, color = mtn_mass_new), alpha = 0.9) +
  ggtitle(i)+scale_color_viridis()
print(x)

# for troubleshooting and checks. 
# plot(mtns$geometry)
# plot(my_centroid, add =T)
# plot(st_make_valid((sf::st_transform(my_centroid_with_b_km_buffer, crs=st_crs(mtns)))), add = T)
# probcoords <- st_geometry(st_sfc(st_point(c(8.80132870725873,18.1810491982364))))
# plot(probcoords, add = T, col = "red")
```


UPDATED CALCS FOR WATER BUFFERING
```{r}
wtr <- read_sf("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/ne_50m_ocean/ne_50m_ocean.shp")
buffer_ranges<-c(seq(from = 0, to = 200, by  = 50)[-1], seq(from=0, to = 3000, by = 250)[-1]) * 1000
bufferdataw <- matrix(nrow = nrow(mydata), ncol = length(buffer_ranges))
for(i in 1:nrow(mydata)){
  my_centroid <- st_geometry(st_sfc(st_point(c(mydata$lon[i],mydata$lat[i]))))
  st_crs(my_centroid) <- crs(LonLat_BirdPAM_raster)
  EPSG_2_UTM <- as.numeric(lonlat2UTM(my_centroid[[1]]))
  # To see the UTM #st_crs(EPSG_2_UTM)$proj4string
  my_centroid_proj = st_transform(st_as_sf(my_centroid), EPSG_2_UTM)
  for(j in 1:length(buffer_ranges)){
    my_centroid_with_b_km_buffer <- st_buffer(my_centroid_proj,dist=buffer_ranges[j])
    wtr_in_buffer_b_km <- st_intersection(st_make_valid(sf::st_transform(my_centroid_with_b_km_buffer, crs=st_crs(wtr))), st_make_valid(wtr))
    area_in_buffer_b_km <- sum(as.numeric(st_area(wtr_in_buffer_b_km)))
    bufferdataw[i, j] <- area_in_buffer_b_km
  }
  print(i)
}
bufferdatawhold <- bufferdataw
for(i in 1:nrow(bufferdataw)) {
  bufferdataw[i,] <- bufferdataw[i,] / buffer_ranges
}
mydata$wtr_mass_new <- rowMeans(bufferdataw)
plot(scale(mydata$wtr_mass_new)~scale(mydata$water_buffering))
abline(0,1)


plotdata <- mydata[order(mydata[, "wtr_mass_new"], decreasing = F),]
x <- ggplot(world)+
  geom_sf() +
  geom_point(data = plotdata[which(plotdata$wtr_mass_new < 20000000),], aes(y=lat, x=lon, color = wtr_mass_new), alpha = 0.9) +
  ggtitle(i)+scale_color_viridis()
print(x)


plot(sort(mydata$wtr_mass_new))
head(mydata[order(mydata$wtr_mass_new, decreasing = T),])

ggplot(mydata[mydata$wtr_mass_new < 20000000,],aes(x=wtr_mass_new, y = mtn_mass_new, color =realm))+
  geom_point()
ggplot(mydata[mydata$wtr_mass_new < 20000000,],aes(x=wtr_mass_new, y = mtn_mass_new, color = vlog(cost)))+
  geom_point()+
  scale_color_viridis()
```

```{r}
mydata[mydata$wtr_mass_new == max(mydata$wtr_mass_new),]
```








Neighbors for sptial analysis. 
```{r}
distm <- mydata[,c("lon", "lat")]
distm <- geodist(distm, measure = "geodesic")
rownames(distm) <- rownames(mydata)
colnames(distm) <- rownames(mydata)
basecols <- ncol(mydata)

# neightbors
coords<-cbind(mydata$lon, mydata$lat); coords<-as.matrix(coords) ; row.names(coords)<-rownames(mydata)
k1 <- knn2nb(knearneigh(coords, longlat = T))
nb<- dnearneigh(coords,row.names = row.names(coords), d1=0,d2=max(unlist(nbdists(k1, coords, longlat = T))),longlat=T)
```


plots for transformations
```{r}
for (i in c("cost", "lat", "lon","tas_breadth","tas_position","pcp_breadth","pcp_position", "mtn_mass", "dispersal_ability", "pair_age", "distance", "boundary_length", "MAT_overlap")) {
hist(mydata[, i], breaks = 50, main = i)
}
```

world plots
```{r}
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Analyze_Processed_Cluster_Outputs/Data")
# pdf(file="variable_maps.pdf", width = 10, height = 7)
for (i in c("cost", "lat", "lon","tas_breadth","tas_position","pcp_breadth","pcp_position", "mtn_mass",  "water_buffering", "dispersal_ability", "pair_age", "distance", "boundary_length", "MAT_overlap", "realm", "landgap")) {
  if(i %in% c("cost")){
    myc <- mydata[, i]; myc <- vlog(myc)
    x <- ggplot(world)+
      geom_sf() +
      geom_point(data = mydata[order(mydata[, i], decreasing = F),], aes(y=lat, x=lon, color = myc), alpha = 0.85) +
      ggtitle(paste0("ln ",i))+
      scale_color_viridis()
    print(x)
  } else if (i %in% c("realm", "landgap")){
    x <- ggplot(world)+
      geom_sf() + 
      geom_point(data = mydata[order(mydata[, "cost"], decreasing = F),], aes(y=lat, x=lon, color = get(i)), alpha = 0.85) +
      ggtitle(paste0(i))+
      scale_color_viridis(discrete = T)
    print(x) 
  } else {
    x <- ggplot(world)+
      geom_sf() + 
      geom_point(data = mydata[order(mydata[, "cost"], decreasing = F),], aes(y=lat, x=lon, color = get(i)), alpha = 0.85) +
      ggtitle(i)+
      scale_color_viridis()
    print(x) 
  }
}
# dev.off()
```


```{r}
mydata <- mydata[, 1:basecols]
# no spatial filtering ---------------------------------------------------------
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML");summary(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)

# plot ---------------------------------
plot(scale(I(vlog(mydata$tas_breadth))), scale(I(vlog(mydata$cost))), pch = 16,
     xlab = "thermal niche breadth", ylab = paste0("log ", costvar, " cost" ), main = "Global model")
myx <- seq(min(scale(vlog(mydata$tas_breadth))), max(scale(vlog(mydata$tas_breadth))), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red')
m <- gls(scale(I(vlog(cost))) ~ scale(I(log(tas_breadth))), data = mydata, method = "REML")
myx <- seq(min(scale(vlog(mydata$tas_breadth))), max(scale(vlog(mydata$tas_breadth))), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red', lty=2)




# with spatial filtering -------------------------------------------------------
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML")
# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # no evidence of spatial component.. 

# spatial filtering --------------------
rm(sarcol)
sarcol <- SpatialFiltering(formula = scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))),
                           data = mydata,nb=nb, style="W", ExactEV = TRUE)
mydata[,c((basecols+1):(basecols+1+ dim(fitted(sarcol))[2]-1))]<-fitted(sarcol)
colnames(mydata)

# vector 1 added ---
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))) + V22 + V23 + V24 + V25 + V26 + V27 + V28 + V29 + V30, correlation = corPagel(0.99, phy = tree,fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m); cor(predict(m),scale(I(vlog(mydata$cost))))
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # too much.
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)

for(i in c("V22", "V23", "V24", "V25", "V26", "V27", "V28", "V29", "V30")){
  x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[,i]),], aes(y=lat, x=lon, color = mydata[,i]), alpha = 0.9) +
  scale_color_viridis()+
  ggtitle(i)
  print(x)
}
```


Is thermal niche breadth predicted by latitude plus latitude2? 
```{r}
mydata <- mydata[, 1:basecols]
m <- gls(scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$tas_breadth))))))
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)

# plot ---------------------------------
plot(scale(mydata$lat), scale(I(vlog(mydata$tas_breadth))), pch = 16,
     xlab = "latitude", ylab = "thermal niche breadth", main = "Global model")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red')
m <- gls(scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)), data = mydata, method = "REML")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red', lty=2)






# with spatial filtering -------------------------------------------------------
m <- gls(scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML")
# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # no evidence of spatial component.. 

# spatial filtering --------------------
rm(sarcol)
sarcol <- SpatialFiltering(formula = scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)),
                           data = mydata,nb=nb, style="W", ExactEV = TRUE)
mydata[,c((basecols+1):(basecols+1+ dim(fitted(sarcol))[2]-1))]<-fitted(sarcol)
colnames(mydata) 

# vector 1 added ---
m <- gls(scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)) + V22 + V23 + V24 + V25 + V26 + V27 + V28 + V29 + V30 + V31 + V32 + V33 + V34 + V35 + V36 + V37, correlation = corPagel(0.99, phy = tree,fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m); cor(predict(m),scale(I(vlog(mydata$cost))))
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # too much.
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)

for(i in c("V22", "V23", "V24", "V25", "V26", "V27", "V28", "V29", "V30", "V31", "V32", "V33", "V34", "V35", "V36", "V37")){
  x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[,i]),], aes(y=lat, x=lon, color = mydata[,i]), alpha = 0.9) +
  scale_color_viridis()
  print(x)
}
```

Is cost  predicted by latitude plus latitude2? 
```{r}
mydata <- mydata[, 1:basecols]
m <- gls(scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)

# plot ---------------------------------
plot(scale(mydata$lat), scale(I(vlog(mydata$cost))), pch = 16,
     xlab = "latitude", ylab = "cost", main = "Global model")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red')
m <- gls(scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)), data = mydata, method = "REML")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red', lty=2)






# with spatial filtering -------------------------------------------------------
m <- gls(scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML")
# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # no evidence of spatial component.. 

# spatial filtering --------------------
rm(sarcol)
sarcol <- SpatialFiltering(formula = scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)),
                           data = mydata,nb=nb, style="W", ExactEV = TRUE)
mydata[,c((basecols+1):(basecols+1+ dim(fitted(sarcol))[2]-1))]<-fitted(sarcol)
colnames(mydata) 

# vector 1 added ---
m <- gls(scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)) + V22 + V23 + V24 + V25, correlation = corPagel(0.99, phy = tree,fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m); cor(predict(m),scale(I(vlog(mydata$cost))))
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # too much.
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)

for(i in c("V22", "V23", "V24", "V25")){
  x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[,i]),], aes(y=lat, x=lon, color = mydata[,i]), alpha = 0.9) +
  scale_color_viridis()
  print(x)
}
```


```{r}
mydata <- mydata[, 1:basecols]
m <- gls(scale(I(vlog(cost))) ~ scale(I((tas_breadth))) + scale(I((tas_position))) + scale(I((pcp_breadth))) + scale(I((pcp_position))) + 
           scale(I((mtn_mass))) + scale(I((water_buffering))) + scale(I((dispersal_ability))) + scale(I((pair_age))) + scale(I((distance))) + scale(I((boundary_length))) + 
           scale(I((MAT_overlap))) + landgap + realm,
         correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML");summary(m)
# car::vif(m) # does not work with realm added in. 
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)


# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # no evidence of spatial component.. 

# spatial filtering --------------------
rm(sarcol)
sarcol <- SpatialFiltering(scale(I(vlog(cost))) ~ scale(I((tas_breadth))) + scale(I((tas_position))) + scale(I((pcp_breadth))) + scale(I((pcp_position))) + 
           scale(I((mtn_mass))) + scale(I((water_buffering))) + scale(I((dispersal_ability))) + scale(I((pair_age))) + scale(I((distance))) + scale(I((boundary_length))) + 
           scale(I((MAT_overlap))) + landgap + realm,
                           data = mydata,nb=nb, style="W", ExactEV = TRUE)
mydata[,c((basecols+1):(basecols+1+ dim(fitted(sarcol))[2]-1))]<-fitted(sarcol)
colnames(mydata) 

# vector 1 added ---
mydata$realm2 <- as.factor(mydata$realm2); mydata$realm2 <- relevel(mydata$realm2, ref = "NT")
m <- gls(scale(I(vlog(cost))) ~ scale(I((tas_breadth))) + scale(I((tas_position))) + scale(I((pcp_breadth))) + scale(I((pcp_position))) + 
           scale(I((mtn_mass))) + scale(I((water_buffering))) + scale(I((dispersal_ability))) + scale(I((pair_age))) + scale(I((distance))) + scale(I((boundary_length))) + 
           scale(I((MAT_overlap))) + landgap + realm + V22 + V23 + V24 + V25 + V26 + V27 + V28 + V29 + V30 + V31 + V32 + V33, correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m); cor(predict(m),scale(I(vlog(mydata$cost))))
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = T)
moran.test(residuals(m), nb2listw(nb))$p.value # too much.
hist(resid(m))
plot(m, resid(., type = "p") ~ fitted(.), abline = 0)
plot(m, scale(I(vlog(cost))) ~ fitted(.), abline = c(0,1))
qqnorm(m)


for(i in c("V22","V23", "V24", "V25", "V26", "V27", "V28", "V29", "V30", "V31", "V32", "V33")){
  x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[,i]),], aes(y=lat, x=lon, color = mydata[,i]), alpha = 0.9) +
  scale_color_viridis()
  print(x)
}
```


Sensitivity Analyses
```{r}
# 1 Pair age (all v. < 8mya (end of uplift of Andes))
# 2 Distance (all v. < 1500*1000) (1500 / 110 = ~ 22 degrees)
# 3 MAT_overlap (> 0% v. > 75% (more restrictive == more conservative for this measure.))
# 4 landgap (all v. nogap) *ALL GAPS ARE < 110km (two water grid cells marked as land for having >50% land @ 0.5 degree resolution.)
```