---
title: "JMPH_basic_model_proofs"
output:
  html_document: default
---

Prepare the Workspace
```{r, Prepare the Workspace}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, error = FALSE) # rmd options
rm(list = ls()); invisible(gc()) # cleaning
```
<br><br><br><br><br>

Control Block
```{r, Control Block} 
# f, f, 0.25 looks nice.
allowmigrants <- F # OPTIONS: T, F 
allowsympatry <- F # OPTIONS: T, F
minoverperc <- 0 # remove pairs that do not have elevational overlap
costvar <- "ele"
```
<br><br><br><br><br>

Packages & Prefs
```{r, Packages & Prefs}
options(scipen = 999) # turn off scientific notation
'%notin%' <- Negate('%in%')
require(ggplot2) # load packages
require(GGally)
require(viridis)
require(caper)
library(dplyr)
library(stringr)
library(maps)
library(ape)
library(EnvStats)
library(forecast)
require(nlme)
require(geodist)
require(letsR)
require(spdep)
require(spatialreg)
require(rnaturalearth)
require(rnaturalearthdata)
require(rgeos)
require(sf)
require(rgdal)
require(raster)
world <- ne_coastline(scale = "medium", returnclass = "sf")
vlog <- function(x){
   log( x + abs(min( x , na.rm = T)) + 1)
}
```
<br><br><br><br><br>

Load Main Data
```{r, Load Main Data}
# main dataframe ---------------------------------
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
load(file =  "Pair_Barrier_Data_FEB2021.rdata")
mydata <- mypairdata; rm(mypairdata)
rownames(mydata) <- mydata$Species.1
mydatahold <- mydata
```
<br><br><br><br><br>

Impose masks & do calcuations
```{r, Impose masks & do calcuations}

# dependent variable: elevational barrier size ---
mydata$cost <- mydata[, paste0(costvar, "_c25")] 

# data filtering -----------------------
# elevational overlap ---
mydata$ele_overlap <- mydata[,paste0(costvar, "_ov_perc_smrnge")]
mydata <- mydata[mydata$ele_overlap > minoverperc,]

# migration ---
if(allowmigrants == F){
  mydata <- mydata[which(mydata$Migration == 1.0),]
}

# patry
if(allowsympatry == F){
  mydata <- mydata[which(mydata[,paste0(costvar, "_c0")] > 0 ),]
}

# update sort order --------------------
mydata$sortorder <- seq(1:nrow(mydata))

# longitude ----------------------------
mydata$lon <- mydata[,paste0("lon_mean_pair_", costvar, "_c25")]

# latitude ----------------------------
mydata$lat <- mydata[,paste0("lat_mean_pair_", costvar, "_c25")]

# temperature breadth -----------------
mydata$tas_breadth <- mydata$tas_range # mean(mean(sp1 annual tas range -- one value per cell), mean(sp2 annual tas range -- one value per cell))
# mydata$tas_breadth <- rowMeans(mydata[, c("mean_VarT_sp1", "mean_VarT_sp2")])

# mean annual temperature --------------
mydata$tas_position <- mydata$tas_mean # mean(mean(sp1 annual tas mean  -- one value per cell), mean(sp2 annual tas mean  -- one value per cell))

# precipitation breadth --------------- 
mydata$pcp_breadth <- mydata$pcp_range # mean(mean(sp1 annual pcp range  -- one value per cell), mean(sp2 annual pcp range  -- one value per cell))
# mydata$pcp_breadth <- rowMeans(mydata[, c("mean_VarP_sp1", "mean_VarP_sp2")])

# precipitation breadth --------------- 
mydata$pcp_position <- mydata$pcp_mean # mean(mean(sp1 annual pcp mean  -- one value per cell), mean(sp2 annual pcp mean  -- one value per cell))

# distance -----------------------------
mydata$distance <- mydata[,paste0("centroid_distance_",costvar,"_c25")]

# mountain mass ------------------------
mtns <- readOGR(dsn="~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/GMBA", layer="GMBA Mountain Inventory_v1.2-World", verbose = FALSE)
wdPAM <- "/Users/boterolab1/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data"
setwd(wdPAM); load("LonLat_BirdPAM_raster.rdata")
mtns <- rasterize(mtns, LonLat_BirdPAM_raster)
mtns@data@values[!is.na(mtns@data@values)] <- 1 # replace mountain IDs with simple coding. 1 for mountain...
mtns@data@values[is.na(mtns@data@values)] <- 0 # ...0 for no mountain

mydata$mtn_mass <- NA
for (i in 1:nrow(mydata)) {
  coords1<-data.frame(lon=mydata$lon[i], lat=mydata$lat[i]); coordinates(coords1)<-c("lon","lat"); crs(coords1)<-crs(LonLat_BirdPAM_raster) # get coordinates
  z <- extract(mtns, coords1, method="bilinear", buffer = raster::pointDistance(c(0,0), c(0,10), lonlat = T))
  mydata$mtn_mass[i] <- sum(z[[1]]) / length(z[[1]])
}

raster::pointDistance(c(0,0), c(0,8), lonlat = T) # corresponds to a radius of just about 8 degrees.
raster::pointDistance(c(75,75), c(75,(75+8)), lonlat = T) # polar circles are bigger, but not that much bigger. so should be OK.
raster::plot(mtns)
raster::plot(world$geometry, add = T)
plotrix::draw.circle(70, 40, 8, nv = 1000, border = "hotpink", lty = 1, lwd = 1)
plotrix::draw.circle(140, -5, 8, nv = 1000, border = "hotpink", lty = 1, lwd = 1)
plotrix::draw.circle(-80, 10, 8, nv = 1000, border = "hotpink", lty = 1, lwd = 1)
plotrix::draw.circle(-70, -50, 8, nv = 1000, border = "hotpink",lty = 1, lwd = 1)

x <- ggplot(world)+
  geom_sf() +
  geom_point(data = mydata[order(mydata[, "mtn_mass"], decreasing = F),], aes(y=lat, x=lon, color = mtn_mass), alpha = 0.9) +
  ggtitle(i)+scale_color_viridis()
print(x)

# water buffering ----------------------
mydata$water_buffering <- NA

# dispersal ability --------------------
dispab <- read.csv("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/Bird Hand-Wing Index/Dataset HWI 2020-04-10.csv")
mydata$dispersal_ability <- NA
for (i in 1:nrow(mydata)){
  dispab_sp1 <- dispab$HWI[dispab$IUCN.name == mydata$Species.1bl[i]]
  dispab_sp2 <- dispab$HWI[dispab$IUCN.name == mydata$Species.2bl[i]]
  dispab_pair <- mean(c(dispab_sp1, dispab_sp2), na.rm = T)
  mydata$dispersal_ability[i] <- dispab_pair
  rm(dispab_sp1, dispab_sp2, dispab_pair)
}
mydata <- mydata[!is.nan(mydata$dispersal_ability),]

# pair age -----------------------------
mydata$pair_age <- mydata$Pair.age..MY.

# length of boundary -------------------
mydata$boundary_length <- mydata$boundary_length_ele_c25

# retain cols of interest only.
mydata <- mydata[,c("uniquePairId", "Species.1", "Species.2", "Species.1bl", "Species.2bl", "cost", "lat", "lon",
                    "tas_breadth","tas_position", "pcp_breadth","pcp_position", "mtn_mass", "water_buffering", 
                    "dispersal_ability", "pair_age", "distance", "boundary_length")]
```

Remove pairs that slipped through the cracks.
```{r}
# # Basic range maps for all pairs (no paths)
# wdPAM <- "/Users/boterolab1/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/PREP/PAM/Data"
# setwd(wdPAM); load("cbPAM.rdata")
# setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Process_Cluster_Outputs/Data")
# pdf("pairmaps.pdf", width = 19, height = 9.25)
# for (i in 1:nrow(mydata)){
#   x <- cbPAM[,c("Longitude(x)","Latitude(y)",mydata$Species.1bl[i])]
# x <- as.data.frame(x[x[,3] == 1,]); colnames(x) <- c("lon", "lat", "pres")
# 
# y <- cbPAM[,c("Longitude(x)","Latitude(y)",mydata$Species.2bl[i])]
# y <- as.data.frame(y[y[,3] == 1,]); colnames(y) <- c("lon", "lat", "pres")
# 
# z <- ggplot(world)+
#     geom_sf() + 
#     geom_point(data = x, aes(y=lat, x=lon), color = "red") + 
#     geom_point(data = y, aes(y=lat, x=lon), color = "green") +
#     theme_bw() + 
#     ggtitle(i)
# print(z)
# }
# dev.off()

# this is for a 253 row mydata, saved as exclusion_mydata 
exclusion <- read.csv("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/Pair_exclusion_by_eye.csv")
exclusion$realm1red[is.na(exclusion$realm1red)] <- "NA"
exclusion$realm2green[is.na(exclusion$realm2green)] <- "NA"

mydata_exclusion <- cbind(mydata, exclusion)
save(mydata_exclusion, file = "mydata_exclusion.rdata")
mydata$realm1 <- exclusion$realm1red
mydata$realm2 <- exclusion$realm2green
mydata$landgap <- exclusion$island
mydata <- mydata[which(exclusion$cosmopolitan == 0 & exclusion$new.old == 0),]

hist(mydata$distance/1000, breaks = 100)
abline(v=2500)
abline(v=quantile(mydata$distance, 0.8)/1000, col = "red")
summary(mydata$distance/(1000))
mydata <- mydata[which(mydata$distance <= (2500*1000)),]

mydata <- mydata[mydata$realm1 == mydata$realm2,]

hist(mydata$pair_age, breaks = 100)
abline(v=8)
mydata <- mydata[which(mydata$pair_age < 7),]
```



Neighbors for sptial analysis. 
```{r}
distm <- mydata[,c("lon", "lat")]
distm <- geodist(distm, measure = "geodesic")
rownames(distm) <- rownames(mydata)
colnames(distm) <- rownames(mydata)
basecols <- ncol(mydata)

# neightbors
coords<-cbind(mydata$lon, mydata$lat); coords<-as.matrix(coords) ; row.names(coords)<-rownames(mydata)
k1 <- knn2nb(knearneigh(coords, longlat = T))
nb<- dnearneigh(coords,row.names = row.names(coords), d1=0,d2=max(unlist(nbdists(k1, coords, longlat = T))),longlat=T)
```
<br><br><br><br><br>

load phylo and prune
```{r, load phylo and prune}
# phylo ------------------------------------------
setwd("~/Box Sync/CB_VF_Shared/Dry_Lab/Projects/JMPH/Other_Input_Data/BirdTrees")
load(file = "BirdTrees.Rdata")
tree <- trees[[1]]; rm(trees) # pick tree (VF GET TREES FROM COONEY!!! and use MCC tree.)
tree <- drop.tip(tree, tree$tip.label[which(tree$tip.label %notin% mydata$Species.1)]) # initial name matching.
mydata <- mydata[which(mydata$Species.1 %in% tree$tip.label),]
mydata <- mydata[match(tree$tip.label, mydata$Species.1),]
sum(mydata$Species.1 != tree$tip.label) # sorted (but still specify form below for safety)
```
<br><br><br><br><br>

plots for transformations
```{r, world plots}
for (i in c("cost", "lat", "lon","tas_breadth","tas_position","pcp_breadth","pcp_position", "mtn_mass", "dispersal_ability", "pair_age", "distance", "boundary_length")) {
hist(mydata[, i], breaks = 50, main = i)
}
```


world plots
```{r, world plots}
for (i in c("cost", "lat", "lon","tas_breadth","tas_position", "pcp_breadth","pcp_position", "mtn_mass", "dispersal_ability", "pair_age", "distance", "boundary_length")) {
  if(i %in% c("cost", "tas_breadth", "mtn_mass", "dispersal_ability", "pair_age", "boundary_length","pcp_breadth","pcp_position")){
  myc <- mydata[, i]; myc <- vlog(myc)
  hist(myc, main = i, breaks = 50)
  } else {
  myc <- mydata[, i];
  hist(myc, main = i, breaks = 50)
  }
}
```
<br><br><br><br><br>

world plots
```{r, world plots}
mydata$realmcolor <- NA
mydata$realmcolor[mydata$realm1 == "NA"] <- "blue"
mydata$realmcolor[mydata$realm1 == "NT"] <- "cyan"
mydata$realmcolor[mydata$realm1 == "PA"] <- "darkred"
mydata$realmcolor[mydata$realm1 == "AT"] <- "darkorange"
mydata$realmcolor[mydata$realm1 == "IM"] <- "red"
mydata$realmcolor[mydata$realm1 == "AA"] <- "hotpink"
mydata$realmcolor[mydata$realm1 == "OC"] <- "grey"

mydata$newoldcolor <- "red"
mydata$newoldcolor[mydata$realm %in% c( "NA", "NT")] <- "blue"

mydata$northsouthcolor <- "green"
mydata$northsouthcolor[mydata$lat >= 0] <- "purple"

mycolor <- mydata$realmcolor

plot(cost ~ lat, col = mycolor,pch=19, data = mydata)
plot(cost ~ abs(lat), col = mycolor,pch=19, data = mydata)
plot(vlog(cost) ~ lat, col = mycolor,pch=19, data = mydata)
plot(vlog(cost) ~ abs(lat), col = mycolor,pch=19, data = mydata)

plot(scale(I(vlog(cost))) ~ scale(I((lat))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I((tas_position))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(pcp_breadth))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(pcp_position))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(mtn_mass))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(dispersal_ability))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(pair_age))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I((distance))), col = mycolor,pch=19, data = mydata)
plot(scale(I(vlog(cost))) ~ scale(I(vlog(boundary_length))), col = mycolor,pch=19, data = mydata)
```

world plots
```{r, world plots}
for (i in c("cost", "lat", "lon","tas_breadth","tas_position", "pcp_breadth","pcp_position", "mtn_mass", "dispersal_ability", "pair_age", "distance", "boundary_length")) {
  if(i %in% c("cost")){
  myc <- mydata[, i]; myc <- vlog(myc)
    x <- ggplot(world)+
    geom_sf() +
    geom_point(data = mydata[order(mydata[, i], decreasing = F),], aes(y=lat, x=lon, color = myc), alpha = 0.9) +
    ggtitle(i)+
    scale_color_viridis()
  print(x)
  } else {
      x <- ggplot(world)+
    geom_sf() + 
    geom_point(data = mydata[order(mydata[, i], decreasing = F),], aes(y=lat, x=lon, color = get(i)), alpha = 0.9) +
    ggtitle(i)+
    scale_color_viridis()
  print(x) 
  }
}

x <- ggplot(world)+
  geom_sf() + 
  geom_point(data = mydata[order(mydata[, "cost"], decreasing = F),], aes(y=lat, x=lon, color = realmcolor), alpha = 0.9) +
  ggtitle(i)
print(x) 
```
<br><br><br><br><br>

Is elevational barrier size predicted by thermal niche breadth?
```{r, Is elevational barrier size predicted by thermal niche breadth}
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth)))*realm1, correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML");summary(m)
anova(m)

m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML");summary(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))
# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = F)
print(paste("moran's I: ", moran.test(residuals(m), nb2listw(nb))$p.value)) # no evidence for spatial autocorrelation.

# plot ---------------------------------
plot(scale(I(vlog(mydata$tas_breadth))), scale(I(vlog(mydata$cost))), pch = 16, col = mydata$realmcolor,
     xlab = "thermal niche breadth", ylab = paste0("log ", costvar, " cost" ), main = "Global model")
myx <- seq(min(scale(vlog(mydata$tas_breadth))), max(scale(vlog(mydata$tas_breadth))), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red')
m <- gls(scale(I(vlog(cost))) ~ scale(I(log(tas_breadth))), data = mydata, method = "REML")
myx <- seq(min(scale(vlog(mydata$tas_breadth))), max(scale(vlog(mydata$tas_breadth))), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red', lty=2)

table(mydata$realm)
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata[mydata$lat >= 0,], method = "REML");summary(m) # there is a not much of a realtionship in the northern hemisphere.

m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata[mydata$lat < 0,], method = "REML");summary(m) # there is a stronger but opposite realtionship in the southern hemisphere.
```
<br><br><br><br><br>

Is thermal niche breadth predicted by latitude plus latitude2? 
```{r, Is thermal niche breadth predicted by latitude plus latitude2}
m <- gls(scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$tas_breadth))))))

# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = F)
print(paste("moran's I: ", moran.test(residuals(m), nb2listw(nb))$p.value)) # no evidence for spatial autocorrelation.

# plot ---------------------------------
plot(scale(mydata$lat), scale(I(vlog(mydata$tas_breadth))), pch = 16, col = mydata$realmcolor,
     xlab = "latitude", ylab = "thermal niche breadth", main = "Global model")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red')
m <- gls(scale(I(vlog(tas_breadth))) ~ scale(lat) + scale(I(lat^2)), data = mydata, method = "REML")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red', lty=2)
```
<br><br><br><br><br>


Is barrier size predicted by latitude plus latitude2? 
```{r, Is thermal niche breadth predicted by latitude plus latitude2}
m <- gls(scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)), correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))

# spatial autocorr ---------------------
matx <- as.matrix(m$residuals); rownames(matx) <- rownames(mydata)
spac <- lets.correl(x=matx, y=distm, z=12, equidistant = T, plot = F)
print(paste("moran's I: ", moran.test(residuals(m), nb2listw(nb))$p.value)) # no evidence for spatial autocorrelation.

# plot ---------------------------------
plot(scale(mydata$lat), scale(I(vlog(mydata$cost))), pch = 16, col = rgb(0,0,0,0.7),
     xlab = "latitude", ylab = "lcost", main = "Global model")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red')
m <- gls(scale(I(vlog(cost))) ~ scale(lat) + scale(I(lat^2)), data = mydata, method = "REML")
myx <- seq(min(scale(mydata$lat)), max(scale(mydata$lat)), length.out=100) # plot fit
mycoefs<-coef(m)
myy <- mycoefs[1] + mycoefs[2]*myx + mycoefs[3]*myx^2
lines(c(0,0), c(-10,100), lty=2)
lines(myx,myy,col = 'red', lty=2)
```


```{r}
# most basic proof -- looks GOOD!
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))),
         correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m); #car::vif(m) # boundary length
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))


# full model -- looks wonky. relationships flip and change a lot!
m <- gls(scale(I(vlog(cost))) ~ scale(I(vlog(tas_breadth))) + scale(tas_position) + scale(pcp_breadth) + scale(pcp_position) + 
           scale(I(vlog(mtn_mass))) + scale(I(vlog(dispersal_ability))) + scale(I(vlog(pair_age))) + scale(I((distance))) + scale(I(vlog(boundary_length))),
         correlation = corPagel(0.99, phy = tree, fixed = F, form = ~Species.1), data = mydata, method = "REML"); summary(m); car::vif(m)
print(paste("Correlation between data and prediction:  ", cor(predict(m),scale(I(vlog(mydata$cost))))))
```







